Supplementary material to the paper 'A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-10-30 analysis/supplementary-materials/supplementary.Rmd
Wokring on comments from reviewers

@@

## Introduction {-}

This document presents the sensitivity analysis performed for informing the number of simulation runs in the analysis of each site, the analysis of each individual site, and the results of re-dating previously shoreline dated sites. All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU

## Sensitivity analysis {-}

@@

## Site analyses {-}

@@

## Numerical result of re-dating {-}

@@

redat <- left_join(tabdat, sitdat, by = c("site_name" = "name")) %>% 
  arrange(site_name) %>% 
  group_by(site_name) %>% 
  mutate(start95 = plyr::round_any(min(start.y), 10),
         end95 = plyr::round_any(max(end.y), 10)) %>% 
  # Replace minus sign/not minus sign with BCE/CE
  mutate(start.x = ifelse(grepl("^[-]", start.x), 
            gsub("^[-]", "", start.x), 
            gsub("$", " CE", start.x)),
         end.x = ifelse(grepl("^[-]", end.x), 
            gsub("^[-](.*)", "\\1 BCE", end.x), 
            gsub("$", " CE", end.x)),
         end95 = ifelse(grepl("^[-]", end95), 
            gsub("^[-](.*)", "\\1 BCE", end95), 
            gsub("$", " CE", end95))) %>%
  mutate(start95 = ifelse(str_detect(end95, "BCE"),
                          ifelse(grepl("^[-]", start95), 
            gsub("^[-](.*)", "\\1", start95), 
            gsub("$", "", start95)),
            ifelse(grepl("^[-]", start95), 
            gsub("^[-](.*)", "\\1 BCE", start95), 
            gsub("$", "CE", start95)))) %>% 
  mutate(elev = paste0(min_elev, "\u2013", max_elev),
         repdate = paste0(start.x, "\u2013", end.x),
         hdrdate = paste0(start95, "\u2013", end95)) %>% 
  select(site_name, elev, repdate, hdrdate, reference)

redat[match(unique(redat$site_name), redat$site_name),] %>% 
   rowid_to_column("id") %>% 
   kable(caption = "Shoreline dating previously shorline dated sites. The range of the 95\\% HDR is rounded off to the nearest 10 years to reduce spurious precision.",
         col.names = c("", 
                     "Site",
                    "Elevation",
                    "Reported date",
                    "95% HDR",
                    "Reference"),
        booktabs = TRUE,  linesep = c(""), escape = TRUE, longtable = TRUE)

>>>>

# Load simulation results and model selection
load(here("analysis/data/derived_data/06data.RData"))

@@

## Contents

This document presents the sensitivity analysis performed for informing the number of simulation runs in the analysis of each site, the analysis of each individual site, the evaluation of different for models for characterising the results, and the results of re-dating previously shoreline dated sites. All underlying and derived data are available in the online repository for the paper at https://osf.io/7F9SU

## Sensitivity analysis

@@

## Site analyses

@@

## Model comparison and selection

In the figure below a series of standard models have been fit to the simulation results older than 2500 BCE (presented with negative values set to zero in Figure 7A in the main text). The models were fit by means of maximum likelihood estimation. Before fitting all models, the values have been adjusted by adding a constant of 0.001 to avoid zero values for the logistic, log normal, power law and gamma distributions. The models have only been fit to the positive values. This was done both when removing the negative values and when forcing these to zero, before adjusting by 0.001. As the difference between these two solutions was negligible and as the assumption here is that the negative values in actuality reflect a distance of zero (see also main text), the result of employing the latter approach is reported on here. The performance of the models were then compared by means of the Akaike information criterion (AIC) and the Bayesian (or Schwarz) information criterion (BIC). The AIC and BIC evaluate the degree to which the models fit to the data, while penalising for the number of model parameters to avoid over-fitting [e.g @burnham2002; for applications in archaeology see e.g. @eve2014; @timpson2021]. As lower values point to a better model, it is evident from the results that the gamma distribution is the best candidate. The approach taken here could have benefited from a more sophisticated treatment of the zero-values, which is related to the fact that these are likely to be a mix of both exact zeros, the case when there is an actual intersection between sea and site, and, although probably to a far lesser extent, zeroes that result from the case when the distance between site and sea is below the detection limit due to the employed methods and the resolution of the spatial data [see e.g. @dunn2005; @helsel2005]. In conclusion, however, the gamma represents a reasonable first approximation.

```{r model_selection, echo = FALSE, out.width = "500px", regfloat=TRUE, warning=FALSE}
include_graphics(here("analysis/figures/model_comparison.png"))
```

``` {r}
# Table with model paramters and BIC+AIC scores

mcdf$parameters <- linebreak(mcdf$parameters)

mcdf[,c(1,6, 3, 5, 4)] %>% 
kable("latex", caption = "Model comparison using AIC and BIC. The models are listed in the order of performance.",
         col.names = c("Model",
                    "Parameters",
                    "AIC",
                    "BIC",
                    "Log-liklihood"),
        booktabs = TRUE, escape = FALSE,
        longtable = TRUE, row.names = FALSE) %>% 
        column_spec(2, width = "6cm") %>% 
        kable_styling(latex_options = "striped")
        # kableExtra::kable_styling(latex_options= c("scale_down"))

## Numerical result of re-dating

@@

redat <- left_join(tabdat, sitdat, by = c("site_name" = "name")) %>% 
  arrange(site_name) %>% 
  group_by(site_name) %>% 
  mutate(start95 = plyr::round_any(min(start.y), 10),
         end95 = plyr::round_any(max(end.y), 10)) %>% 
  # Replace minus sign/not minus sign with BCE/CE
  mutate(start.x = ifelse(grepl("^[-]", start.x), 
            gsub("^[-]", "", start.x), 
            gsub("$", " CE", start.x)),
         end.x = ifelse(grepl("^[-]", end.x), 
            gsub("^[-](.*)", "\\1 BCE", end.x), 
            gsub("$", " CE", end.x)),
         end95 = ifelse(grepl("^[-]", end95), 
            gsub("^[-](.*)", "\\1 BCE", end95), 
            gsub("$", " CE", end95))) %>%
  mutate(start95 = ifelse(str_detect(end95, "BCE"),
                          ifelse(grepl("^[-]", start95), 
            gsub("^[-](.*)", "\\1", start95), 
            gsub("$", "", start95)),
            ifelse(grepl("^[-]", start95), 
            gsub("^[-](.*)", "\\1 BCE", start95), 
            gsub("$", "CE", start95)))) %>% 
  mutate(elev = paste0(min_elev, "\u2013", max_elev),
         repdate = paste0(start.x, "\u2013", end.x),
         hdrdate = paste0(start95, "\u2013", end95)) %>%
  dplyr::select(site_name, elev, repdate, hdrdate, reference)

redat[match(unique(redat$site_name), redat$site_name),] %>% 
   rowid_to_column("id") %>% 
   kable(caption = "The result of shoreline dating previously shorline dated sites using the method proposed here. The range of the 95\\% HDR is rounded to the nearest 10 years to reduce spurious precision.",
         col.names = c("",
                     "Site",
                    "Reported elevation",
                    "Reported date",
                    "95% HDR",
                    "Reference"),
        booktabs = TRUE,  linesep = c(""), escape = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = "striped")
